Approximation as Differentiation

### 2025-06-07 main/first-order.tex
FinVect and LatGal as examples of semi-additive categories.

\subsection{Category of Lattices and Galois Connections}

We start with the category of lattices and Galois connections, which will serve as our base category for
interpreting forward and backward \GPS. Galois connections are pairs of monotone functions $f: Y \to X$ and
$g: X \to Y$ between posets, where $f$ is the (pointwise) best approximation from below to an inverse of $g$,
and $g$ the best approximation from above to an inverse of $f$. This setup is a nice fit for the bidirectional
nature of (dynamic) program slicing, with $f$ capturing \emph{backward} slicing (producing the least input
slice for a given slice of the output), and $g$ capturing \emph{forward} slicing (producing the greatest
output slice for a given slice of the input).

@@

The meet-preserving maps from $X$ to $Y$ themselves form a meet-semilattice. The operation $\meet$ on such
maps is given pointwise, so that $(f \meet g)(x) = f(x) \meet g(x): X \to Y$, and preserves meets; the
constant map $\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$, which also preserves meets, provides the
unit. Dually the join-preserving maps from $Y$ to $X$ have a join-semilattice structure given pointwise by
$\join$ and the constant map $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to $\bot_X$. Since these two
constructions come in adjoint pairs, the hom-set $\LatGal(X,Y)$ forms a bounded meet semilattice with unit
$\top_{X,Y}$ given by the Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and meet of Galois
connections $(f \adj g) \meet (f' \adj g') = (f \join f') \adj (g \meet g'): X \to Y$. Thus $\LatGal$ is
enriched in $\SemiLat$, the category of semilattices and semilattice homomorphisms.

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and so $\cat{C}$ has only a single object, up to isomorphism. \todo{Bridge better to
next section; even if $\LatGal$ were Cartesian closed, directly interpreting into $\LatGal$ wouldn't allow
each point to be associated with its own lattice of approximants.}

@@

Following the CHAD approach, automatic differentiation for first-order programs can also be recovered in this
setup, via the total category $\Fam(\FinVect)$. We recapitulate the basic ideas here, starting with the
category $\FinVect$ of finite vector spaces over $\RR$, with addition of real numbers written $a + b$ and
multiplication written $a \mult b$.

\subsubsection{Category of finite-dimensional vector spaces}
\label{sec:categories-with-biproducts:fdvect}

\begin{definition}[Finite dimensional vector space over $\RR$]
Up to isomorphism, an \emph{$n$-dimensional vector space over $\RR$} is the set $\RR^n$ equipped with
component-wise addition of vectors $+: \RR^n \times \RR^n \to \RR^n$ and scalar multiplication $\mult: \RR
\times \RR^n \to \RR^n$ (overloading $+$ and $\mult$), where $(\RR^n,+)$ is an abelian group (with identity
$0$ and additive inverse $-v$ again defined component-wise), and where the vector operations are compatible
with the field operations in that the following equations hold:

\begin{definition}[Category $\FinVect$]
Define $\FinVect$ to be the category which has as objects $\RR^n$ all finite-dimensional vector spaces, and as
morphisms $f: \RR^m \to \RR^n$ all functions $f$ satisfying $f(u + v) = f(u) + f(v)$ and $f(a \mult v) = a
\mult f(v)$, with $f$ preserving the abelian group structure automatically.
\end{definition}

$\FinVect$ is enriched in $\Ab$, the category of abelian groups. For any objects $V = \RR^m$ and $W = \RR^n$
the hom-set $\FinVect(V,W)$ is an abelian group, with identity $0_{V,W}$ given by the constant map $v \mapsto
0_W$ and addition and additive inverse given pointwise:

\vspace{1mm}
\noindent Composition is with bilinear. The direct sum $V \biprod W \iso \RR^{m + n}$ given by the Cartesian
product is a biproduct $(V \biprod W, \biinj_{V}, \biinj_{W}, \pi_1, \pi_2)$ with $\biinj_{V} =
\prodM{\id_{V}}{0_{V,W}}$ and $\biinj_{W} = \prodM{0_{V,W}}{\id_{W}}$. \todo{Significance of biproducts}

>>>>

\subsection{Biproducts and Semi-Additive Categories}
\label{sec:biproducts}

\todo{Introduce $\CMon$-category first.} We start with the idea of a biproduct and semi-additive categories,
which are always enriched in $\CMon$, the category of commutative monoids.

\begin{definition}[Semi-additive category]
\label{def:biproducts:semi-additive}
A category with finite products and coproducts is \emph{semi-additive} if the canonical morphisms (projections
and injections) give an isomorphism
\[\textstyle X \coprod Y \iso X \times Y\] that is natural in both variables.
\end{definition}

The product/coproduct is called a \emph{biproduct}, with the biproduct structure denoted by $(\biprod, 0)$ and
projections $\biproj_X, \biproj_Y$ and injections $\biinj_X$ and $\biinj_Y$. The unit $0$ is both terminal and
initial and is called a \emph{zero} object. A semi-additive category $\cat{C}$ is enriched in $\CMon$: for any
two morphisms $f, g: X \to Y$ in $\cat{C}$, the biproduct structure provides a way to ``add'' them together,
forming a morphism $f + g: X \to Y$. Diagrammatically:

\begin{center}
\begin{tikzcd}
   X \arrow[r, "\diag"] & X \biprod X \arrow[r, "f \biprod g"] & Y \biprod Y \arrow[r, "\codiag"] & Y
\end{tikzcd}
\end{center}

Here $\diag$ denotes the diagonal $\prodM{\id_X}{\id_X}$ given by the universal property of the product and
$\codiag$ denotes the codiagonal $\coprodM{\id_X}{\id_Y}$ given by the universal property of the coproduct.
The $f \oplus g$ morphism is the component-wise map $\prodM{f \comp \biproj_X}{g \comp \biproj_Y} =
\coprodM{\biinj_X \comp f}{\biinj_Y \comp g}$. Similarly we can exhibit a \emph{zero} morphism $0_{X,Y}$ by
composing the unique maps in and out of the zero object:

\begin{center}
\begin{tikzcd}
   X \arrow[r, "!_X"] & 0 \arrow[r, "!^Y"] & Y
\end{tikzcd}
\end{center}

It is easy to verify that $+$ is associative and commutative and that $f + 0_{X,Y} = f$, and thus that every
hom-object $\cat{C}(X,Y)$ is an object in $\CMon$. Moreover composition is \emph{bilinear}, i.e.~given by a
family of morphisms $\Hom{\cat{C}}{Y}{Z} \tensor \Hom{\cat{C}}{X}{Y} \to \Hom{\cat{C}}{X}{Z}$ in $\CMon$ that
preserve the additive structure in $\Hom{\cat{C}}{Y}{Z}$ and $\Hom{\cat{C}}{X}{Y}$ separately:

\begin{salign*}
f \comp \zero_{X,Y} = 0_{X,Z} = \zero_{Y,Z} \comp f
\end{salign*}
\begin{salign*}
(f + g) \comp h &= (f \comp h) + (g \comp h) \\
h \comp (f + g) &= (h \comp f) + (h \comp g)
\end{salign*}

\subsubsection{Biproduct laws}
It is also easy to show that in a semi-additive category the following equations hold:
%\vspace{-4mm}
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_X \comp \biinj_X &= \id_X \\
   \biproj_Y \comp \biinj_X &= \zero_{X,Y}
\end{salign*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_Y \comp \biinj_Y &= \id_Y \\
   \biproj_X \comp \biinj_Y &= \zero_{Y,X}
\end{salign*}
\end{center}
\end{minipage}

\begin{salign*}
(\biinj_X \comp \biproj_X) + (\biinj_Y \comp \biproj_Y) &= \id_{X \biprod Y}
\end{salign*}

\vspace{3mm}
\noindent It is also possible to use these laws to define biproducts, in the case where $\cat{C}$ is enriched
in $\CMon$ and so equipped with addition of morphisms and zero morphisms. In that case any products or
coproducts in $\cat{C}$ are necessarily biproducts.

\begin{proposition}
\label{prop:biproducts:from-product-or-coproduct}
In a category $\cat{C}$ enriched in $\CMon$:
\begin{enumerate}
\item If $(X \times Y, \pi_1, \pi_2)$ is a product then $(X \times Y, \inj_1, \inj_2)$ is a coproduct, where
$\inj_X = \prodM{\id_X}{0_{X,Y}}: X \to X \times Y$ and $\inj_Y = \prodM{0_{X,Y}}{\id_Y}: Y \to X \times Y$.
\item If $(X \coprod Y, \inj_1, \inj_2)$ is a coproduct then $(X \coprod Y, \pi_1, \pi_2)$ is a product, where
$\pi_1 = \coprodM{\id_X}{0_{Y,X}}: X \coprod Y \to X$ and $\pi_2 = \coprodM{0_{X,Y}}{\id_Y}: X \coprod Y \to
Y$.
\item If $X$ is terminal (resp.~initial) then $X$ is initial (resp.~terminal).
\end{enumerate}
\end{proposition}

\subsubsection{Example: Category of Finite Vector Spaces}
\label{sec:categories-with-biproducts:fdvect}

The category $\FinVect$ of finite vector spaces over $\RR$ (with addition of real numbers written $a + b$ and
multiplication written $a \mult b$) is semi-additive.

\begin{definition}[Finite dimensional vector space over $\RR$]
Up to isomorphism, an \emph{$n$-dimensional vector space over $\RR$} is the set $\RR^n$ equipped with
component-wise addition of vectors and scalar multiplication (also written $+$ and $\mult$), where $(\RR^n,+)$
is an abelian group (with identity $0$ and additive inverse $-v$ again defined component-wise), and where the
vector operations are compatible with the field operations in that the following equations hold \todo{probably
more detail than we need to go into}:

\vspace{1mm}
\noindent Composition is with bilinear. The direct sum $V \biprod W \iso \RR^{m + n}$ given by the Cartesian
product is a biproduct $(V \biprod W, \biinj_{V}, \biinj_{W}, \pi_1, \pi_2)$ with $\biinj_{V} =
\prodM{\id_{V}}{0_{V,W}}$ and $\biinj_{W} = \prodM{0_{V,W}}{\id_{W}}$.

\begin{definition}
Define $\FinVect$ to be the category which has as objects $\RR^n$ all finite-dimensional vector spaces, and as
morphisms $f: \RR^m \to \RR^n$ all functions $f$ satisfying $f(u + v) = f(u) + f(v)$ and $f(a \mult v) = a
\mult f(v)$, with $f$ preserving the abelian group structure automatically.
\end{definition}

$\FinVect$ is enriched in $\Ab$, the category of abelian groups \todo{Introduce $\CMon$-enrichment first, then
perhaps $\Ab$-enrichment parenthetically}. For any objects $V = \RR^m$ and $W = \RR^n$ the hom-set
$\FinVect(V,W)$ is an abelian group, with identity $0_{V,W}$ given by the constant map $v \mapsto 0_W$ and
addition and additive inverse given pointwise:

\subsubsection{Example: Category of Lattices and Galois Connections}

\todo{Rework to account for new ordering.} Galois connections are pairs of monotone functions $f: Y \to X$ and
$g: X \to Y$ between posets, where $f$ is the (pointwise) best approximation from below to an inverse of $g$,
and $g$ the best approximation from above to an inverse of $f$. This setup is a nice fit for the bidirectional
nature of (dynamic) program slicing, with $f$ capturing \emph{backward} slicing (producing the least input
slice for a given slice of the output), and $g$ capturing \emph{forward} slicing (producing the greatest
output slice for a given slice of the input).

@@

\todo{Again, introduce $\CMon$-enrichment first and relegate $\SemiLat$-enrichment to parenthetical remark.}
The meet-preserving maps from $X$ to $Y$ themselves form a meet-semilattice. The operation $\meet$ on such
maps is given pointwise, so that $(f \meet g)(x) = f(x) \meet g(x): X \to Y$, and preserves meets; the
constant map $\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$, which also preserves meets, provides the
unit. Dually the join-preserving maps from $Y$ to $X$ have a join-semilattice structure given pointwise by
$\join$ and the constant map $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to $\bot_X$. Since these two
constructions come in adjoint pairs, the hom-set $\LatGal(X,Y)$ forms a bounded meet semilattice with unit
$\top_{X,Y}$ given by the Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and meet of Galois
connections $(f \adj g) \meet (f' \adj g') = (f \join f') \adj (g \meet g'): X \to Y$. Thus $\LatGal$ is
enriched in $\SemiLat$, the category of semilattices and semilattice homomorphisms.

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and so $\cat{C}$ has only a single object, up to isomorphism.

@@

Following the CHAD approach, automatic differentiation for first-order programs can also be recovered in this
setup, via the total category $\Fam(\FinVect)$. We recapitulate the basic ideas here.
